D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2023-12-22 ootavacuous.tex
Add invented-load litmus test and commentary
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Listing~\ref{lst:Inventing Atomic Loads}
and~\ref{lst:Atomic Loads Invented}
shows a TBD point about inventing atomic loads.

@@

@@

>>>>

This section contains several example litmus tests from the active
mind of Alan Stern.

@@

First, consider Listing~\ref{lst:Inventing Atomic Loads}.
At first glance, this might appear straightforward.
\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
If both dependencies are semantic dependency, the OOTA cycle cannot
form.

@@

But is \co{P0()}'s dependency a semantic dependency?
For example, is a real-world C++ implementation permitted to
duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
values?
If this is permitted, then the implementation would be within its rights
to transform
Listing~\ref{lst:Inventing Atomic Loads}
into
Listing~\ref{lst:Atomic Loads Invented}.

Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
either zero or one, if the values returned by the two loads from \co{x}
differ, then one of those two values must be equal to that from \co{y}.
If the implementation is further permitted to make a runtime choice
between the two values loaded from \co{x}, it could always choose the
one that was equal to the value loaded from \co{y}, as shown by the
\co{if} statement on line~13.
This would mean that the store to \co{z} was always the constant one,
which would permit the store to be ordered before the load from \co{y},
as shown in lines~14 and~15.
Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.

The implementation might also drop the load from \co{y}, whose value
is now unused except by the \co{exists} clause.
However, we use the convention that the \co{exists} clause is observable
behavior.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-unused-load.litmus @@
\caption{Unused Extra Atomic Load}
\label{lst:Unused Extra Atomic Load}
\end{listing}

One possible reaction to this situation is to declare that useful
real-world C++ implementations should never duplicate non-volatile
relaxed atomic loads.
But suppose that the initial code was as shown in
Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
	Such unused loads can easily be generated by complex macros or
	template metaprograms.}
In this case, the transformation to
Listing~\ref{lst:Atomic Loads Invented}
would not require inventing a load, but instead merely making use of
the previously discarded value in \co{r0}.

There are a number of possible reactions to this situation:

\begin{enumerate}
\item	Attempts to use real-world constraints to prevent OOTA cycles
	are futile.
	What to do instead?
\item	Implementations should not invent uses for values from
	non-volatile relaxed atomic loads.
	But are there any useful optimizations that invent such uses?
	(One can argue that in all such cases, the loads should instead
	be fused into a single load.)
\item	Listing~\ref{lst:Inventing Atomic Loads}
	looks like it has an OOTA cycle, but there is in fact no
	semantic dependency.
	But then how exactly does one identify a semantic dependency?
\item	What other equally interesting situations might come to light?
\item	Your reaction here!
\end{enumerate}
